\documentclass[11pt]{formatting-template}

% mathptmx is a Times Roman look-alike (don't use the times package)
% It isn't clear if Times is required. The OGS manual lists several
% "standard fonts" but never says they need to be used.
% \usepackage{mathptmx}

\usepackage[NoDate]{currvita}
\usepackage{array}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}
\usepackage{microtype}
\usepackage[breaklinks=true,pdfborder={0 0 0}]{hyperref}
\usepackage{graphicx}
\AtBeginDocument{%
	\settowidth\cvlabelwidth{\cvlabelfont 0000--0000}
}

% OGS recommends increasing the margins slightly.
\increasemargins{.1in}

\usepackage[english]{babel}
\usepackage{blindtext}

\usepackage{url}

\usepackage[super,sort&compress]{natbib}

\overfullrule5pt

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Thesis Title & Author Information
\title{Interpreting the regulatory genome through predictive modeling}
\author{Adam Klie}
\degree{Bioinformatics \& Systems Biology}{Doctor of Philosophy}

% Thesis Committee Members
\chair{Professor Hannah Carter}
\cochair{Professor Emma Farley} % alphabetical order required
\committee{Professor Gary Cottrell}
\committee{Professor Kyle J. Gaulton}
\committee{Professor Jill Mesirov}
\degreeyear{2025}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

% Begin with front matter and so forth
\frontmatter
\maketitle{}
\makecopyright{}
\makesignature{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dedication
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{dedication}
	\setsinglespacing{}
	\parindent0pt\parskip\baselineskip{}
	\vskip0pt plus.25fil
	\begin{center}

		This dissertation is dedicated to my family, friends, coaches and colleagues,\\
		from whom I owe every ounce of strength I have.

		To my mother, Mary Ann Tomaszewski,\\ 
		for her unwavering support.\\
		Thank you for always being just a phone call away and getting me through the tough times.

		To my father, Ross Klie,\\ 
		for forever being my rebounder, jam buddy, poker partner, caddy and so much more.\\
		Remember, ‘the blues don’t change’.

		To my brother, Nathaniel Klie,\\ 
		for teaching me how to take my lumps on the chin and score the bucket anyway.\\ 
		I would not be the person I am today without you.

		Peace out, one love, thumbs up, hugs.

	\end{center}

\end{dedication}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Epigraph
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{epigraph}
	\vskip0pt plus.5fil
	\vfil\vfil
	\setsinglespacing{}
	{ \flushright{
			I am happy because I want nothing from anyone.\\
			I do not care about money.\\
			Decorations, titles or distinctions mean nothing to me.\\
			I do not crave praise.\\
			The only thing that gives me pleasure, apart from my work, my violin, and my sailboat,\\
			is the appreciation of my fellow workers.\\
			}
			\vskip\baselineskip{}
			- \textit{Albert Einstein}\par
	}
	\vfil\vfil\vfil
\end{epigraph}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Contents, Figures, Tables, Etc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Next comes the table of contents, list of figures, list of tables,
% etc. If you have code listings, you can use \listoflistings (or
% \lstlistoflistings) to have it be produced here as well. Same with
% \listofalgorithms.

\tableofcontents
\listoffigures
\listoftables

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}

	The content of Chapter 1, in full, is a reformatted reprint of the material as it appears in “Predictive analyses of regulatory sequences with EUGENe" in \textit{Nature Computational Science}, 2023 by Adam Klie, David Laub, James V. Talwar, Hayden Stites, Tobias Jores, Joe J. Solvason, Emma K. Farley and Hannah Carter. The dissertation author was the primary investigator and author of this paper.

	The content of Chapter 2, in full, is a reformatted reprint of the material currently being prepared for submission for publication as “Syntax-driven design of neural-specific enhancers in vivo'' by Adam Klie, Krissie Tellez, Joe J Solvason, Fabian Lim, Simran K Jandu, Joseph Hwang, Emma K Farley, and Hannah Carter. The dissertation author was the primary investigator and author of this paper.

	The content of Chapter 3, in full, is a reformatted reprint of the material currently being prepared for submission for publication as “Unraveling the sequence determinants of environment-induced changes in pancreatic beta-cell states'' by Adam Klie, Mei-lin Okino, Hannah M Mummey, Nathan R Zemke, Shivani Lakkaraju, Weronika M Bartosik, Paola Benaglio, Gaowei Wang, Kai Zhang, Han Zhu, Bing Ren, Maike Sander, Kyle J Gaulton, and Hannah Carter. The dissertation author was the primary investigator and author of this paper.

\end{acknowledgements}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Vita
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{vita}
\noindent
\begin{cv}{}
\begin{cvlist}{}
	\item[2017] University of California San Diego\\
		\textit{Bachelor of Science, Bioengineering: Bioinformatics}
	\item[2025] University of California San Diego\\
		\textit{Doctor of Philosophy, Bioinformatics and Systems Biology}
\end{cvlist}
\end{cv}

%%%%% PUBLICATIONS %%%%%
\noindent \textit{Author names marked with $\dagger$ indicate shared first co-authorship.} \newline
\noindent \textit{Publications marked with $\triangle$ are included in this text.} \newline

\noindent \textbf{Adam Klie}, Krissie Tellez, Joe J. Solvason, Simran K. Jandu, Fabian Lim, Joseph Hwang, Emma K. Farley, Hannah Carter. ``Syntax-driven design of neural-specific enhancers in vivo.'', \textit{In Preparation}, 2025. $\triangle$ \newline

\noindent \textbf{Adam Klie}, Mei-lin Okino, Hannah M. Mummey, Nathan R. Zemke, Shivani Lakkaraju, Weronika M. Bartosik, Paola Benaglio, Gaowei Wang, Kai Zhang, Han Zhu, Bing Ren, Maike Sander, Kyle J. Gaulton, Hannah Carter. ``Unraveling the sequence determinants of environment-induced changes in pancreatic beta cell states.'', \textit{In Preparation}, 2025. $\triangle$ \newline

\noindent Juan Carlos Jado, Michelle Dow, Krypton Carolino, \textbf{Adam Klie}, Gregory J. Fonseca, Trey Ideker, Hannah Carter, Elizabeth A. Winzeler. ``In vitro evolution and whole genome analysis to study chemotherapy drug resistance in haploid human cells.'', \textit{Scientific Reports}, 2024. \newline

\noindent Caroline S. Jansen, Meghana S. Pagadala, Maria A. Cardenas, Roshan S. Prabhu, Subir Goyal, Chengjing Zhou, Prasanthi Chappa, Baohan T. Vo, Chengyu Ye, Benjamin Hopkins, Jim Zhong, \textbf{Adam Klie}. ``Pre-operative stereotactic radiosurgery and peri-operative dexamethasone for resectable brain metastases: a two-arm pilot study evaluating clinical outcomes and immunological correlates.'', \textit{Nature Communications}, 2024. \newline

\noindent James V. Talwar, \textbf{Adam Klie}, Meghana S. Pagadala, Hannah Carter. ``GRIEVOUS: your command-line general for resolving cross-dataset genotype inconsistencies.'', \textit{Bioinformatics}, 2024. \newline

\noindent Granton A. Jindal, Alexis T. Bantle, Joe J. Solvason, Jessica L. Grudzien, Agnieszka D’Antonio-Chronowska, Fabian Lim, Sophia H. Le, et al. ``Single-nucleotide variants within heart enhancers increase binding affinity and disrupt heart development.'', \textit{Developmental Cell}, 2023. \newline

\noindent \textbf{Adam Klie}, David Laub, James V. Talwar, Hayden Stites, Tobias Jores, Joe J. Solvason, Emma K. Farley, Hannah Carter. ``Predictive analyses of regulatory sequences with EUGENe.'', \textit{Nature Computational Science}, 2023. $\triangle$ \newline

\noindent \textbf{Adam Klie}$\dagger$, Brian Y. Tsui$\dagger$, Shamim Mollah, Dylan Skola, Michelle Dow, Chun-Nan Hsu, Hannah Carter. ``Increasing metadata coverage of SRA BioSample entries using deep learning-based named entity recognition.'', \textit{Database: The Journal of Biological Databases and Curation}, 2021. \newline

\noindent Sally L. Baxter, \textbf{Adam R. Klie}, Bharanidharan Radha Saseendrakumar, Gordon Y. Ye, Michael Hogarth. ``Text processing for detection of fungal ocular involvement in critical care patients: cross-sectional study.'', \textit{Journal of Medical Internet Research}, 2020. \newline

%%%%% FIELDS OF STUDY %%%%% 
% This section was omitted within this thesis as it was optional.

\end{vita}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dissertation Abstract 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{dissertationabstract}

		The completion of the Human Genome Project in 2003 provided a nearly complete map of human DNA, laying the foundation for the field of genomics. Today, we have genome sequences for over half a million individuals and have identified hundreds of thousands of genetic variants associated with common diseases, most of which lie in the non-coding genome. Despite major advances in biochemical and functional annotation across diverse cell types, the biological mechanisms linking non-coding genetic variation to function remain poorly understood.

		In this thesis, I use predictive modeling to analyze large-scale genomics datasets, with the goal of uncovering how non-coding DNA influences gene regulation across cell types and states. While machine learning has transformed fields such as language and vision, its application to genomics poses distinct challenges. To address these, I present EUGENe (Elucidating the Utility of Genomic Elements with Neural Nets), a flexible, user-friendly software ecosystem that enables the development, training, and interpretation of predictive models for genomic data. EUGENe includes several modular tools designed to be scalable, adaptable, and accessible to a wide range of users, and represents a step forward in democratizing the use of deep learning in genomics.

		I then apply EUGENe to two biological systems. First, I analyze a high-throughput reporter assay of synthetic DNA to investigate how transcription factor binding site syntax shapes enhancer activity. I develop a model that predicts enhancer function from binding site syntax alone, assess its interpretability and generalizability, and use it to design synthetic, tissue-specific enhancers. Second, I apply EUGENe to single-cell chromatin accessibility profiles from stimulated, stem cell–derived pancreatic organoids. I train accurate sequence-to-profile models across 15 cell states, interpret learned regulatory features, and use the models to predict the effects of non-coding variants, revealing mechanisms underlying environment-induced changes in beta cell gene regulation.

		Together, this thesis demonstrates the power of interpretable deep learning models to decode gene regulatory logic and provides tools and frameworks that can be broadly applied to future functional genomics studies.

\end{dissertationabstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main Text of the Dissertation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\mainmatter{}

\include{0_introduction}
\include{1_chapter}
\include{2_chapter}
\include{3_chapter}
\include{4_epilogue}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendix of the Dissertation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\include{1_chapter_appendix}
\include{2_chapter_appendix}
\include{3_chapter_appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% End of the Dissertation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter{}
\bibliographystyle{unsrtnat} 
\bibliography{bibliography}

\end{document}
